---
title: "TMDL Dashboard Downloadable Report"
output: 
  html_document: default
  pdf_document: default
  word_document: default
params:
  year_min: 1975
  year_max: 2025
  region: !r list("05", "07")
  state: !r list("IL", "MO")
  group_poll: "NUTRIENTS"
  name_poll: NULL
  addparam: NULL
  filt.dt: TRUE
  prod.output.plot: TRUE
  prod.output.data: TRUE
  annual.output.plot: TRUE
  annual.output.data: TRUE
  poll.output.plot: TRUE
  poll.output.data: TRUE
  state.output.plot: TRUE
  state.output.data: TRUE
---

```{r setup, echo = FALSE}
load("EQ_data.RData")

library(dplyr)
library(data.table)
library(plotly)
library(stringi)
```

This report was generated from EPA's internal TMDL Dashboard on `r format(Sys.Date(), "%B %d, %Y")`. The Expert Query data used to generate the following tables and figures was updated on `r update.tmdls`.


```{r params.df, echo=FALSE}

desc.param <- c("Date Range For Fiscal Year Established",
                "EPA Region(s)",
                "State(s)",
                "Pollutant Group(s)",
                "Pollutant(s)",
                "Addressed Parameter(s)")

list.params <- c(paste0(params$year_min, "-", params$year_max),
                stringi::stri_replace_last(paste(params$region, collapse = ", "), 
                                          fixed = ", ", " and "),
                stringi::stri_replace_last(paste(params$state, collapse = ", "), 
                                          fixed = ", ", " and "),
                stringi::stri_replace_last(paste(params$group_poll, collapse = ", "), 
                                          fixed = ", ", " and "),
                stringi::stri_replace_last(paste(params$name_poll, collapse = ", "), 
                                          fixed = ", ", " and "),
                stringi::stri_replace_last(paste(params$addparam, collapse = ", "), 
                                          fixed = ", ", " and "))

param.df <- data.frame(desc.param, list.params) %>%
  dplyr::rename("Data Filter" = desc.param,
                "User Input" = list.params)

DT::datatable(param.df)

```

The following parameters were selected by the user. Explain and and or in the selections.

```{r params.dt, echo = TRUE}

DT::datatable(param.df)
```



```{r tmdl.data, echo = FALSE, eval = TRUE}

# this filter doesn't work when one of the params is null - need to address that
tmdl.df <- filt.df %>%
  dplyr::filter(state %in% params$state,
                fiscalYearEstablished >= params$year_min,
                fiscalYearEstablished <= params$year_max,
                region %in% params$region,
                pollutantGroup %in% params$pollgroup,
                pollutant %in% params$poll,
                addressedParameters %in% params$addparam)

tmdl.actid.count <- dplyr::n_distinct(tmdl.df)

tmdl.noactid.count <- tmdl.df %>%
  dplyr::select(pollutant, assessmentUnitId) %>%
  dplyr::distinct() %>%
  dplyr::n_distinct()
```


In this filtered data set there are: 
`r formatC(tmdl.actid.count, big.mark = ",")` unique combinations of actionId, assessmentUnitId and pollutant.
`r formatC(tmdl.noactid.count, big.mark = ",")` unique combinations of assessmentUnitId and pollutant.

`r  if(params$filt.dt == TRUE){"#Header\nIs this under the header\n\n##subheader\nis this undersub?"}`

```{r filt.df.section}



```

