---
title: "TMDL Dashboard Downloadable Report"
output: 
  html_document: default
  pdf_document: default
  word_document: default
params:
  year_min: 1975
  year_max: 2025
  region: !r list("05", "07")
  state: !r list("IL", "MO")
  group_poll: "NUTRIENTS"
  name_poll: NULL
  addparam: NULL
  filt.dt: TRUE
  prod.output.plot: TRUE
  prod.output.data: TRUE
  annual.output.plot: TRUE
  annual.output.data: TRUE
  poll.output.plot: TRUE
  poll.output.data: TRUE
  state.output.plot: TRUE
  state.output.data: TRUE
---

```{r setup, echo = FALSE}
load("EQ_data.RData")

library(dplyr)
library(data.table)
library(plotly)
library(stringi)
```

## Data and Analyis Source

This report was generated from EPA's internal TMDL Dashboard on `r format(Sys.Date(), "%B %d, %Y")`. The Expert Query data used to generate the following tables and figures was updated on `r update.tmdls`.

```{r params.df, echo=FALSE}

desc.param <- c("Date Range For Fiscal Year Established",
                "EPA Region(s)",
                "State(s)",
                "Pollutant Group(s)",
                "Pollutant(s)",
                "Addressed Parameter(s)")

list.params <- c(paste0(params$year_min, "-", params$year_max),
                stringi::stri_replace_last(paste(params$region, collapse = ", "), 
                                          fixed = ", ", " and "),
                stringi::stri_replace_last(paste(params$state, collapse = ", "), 
                                          fixed = ", ", " and "),
                stringi::stri_replace_last(paste(params$group_poll, collapse = ", "), 
                                          fixed = ", ", " and "),
                stringi::stri_replace_last(paste(params$name_poll, collapse = ", "), 
                                          fixed = ", ", " and "),
                stringi::stri_replace_last(paste(params$addparam, collapse = ", "), 
                                          fixed = ", ", " and "))

param.df <- data.frame(desc.param, list.params) %>%
  dplyr::rename("Data Filter" = desc.param,
                "User Input" = list.params)

```

## User Inputs (Data Filters)

The following parameters were selected by the user. Explain and and or in the selections.


```{r params.dt, echo=FALSE}

DT::datatable(param.df)
```

```{r tmdl.data, include=FALSE}

# function to filter data based on user inputs
paramCheck <- function(.data, paramname = NULL) {
  
  if(!is.null(paramname)) {
    
    # create variable to refer to param
    param.name <- paste0("params$", paramname)
    
    # add crosswalk between param names and data set column names
    tmdl.col <- dplyr::case_when(paramname == "year_min" ~ "fiscalYearEstablished",
                                 paramname == "year_max" ~ "fiscalYearEstablished",
                                 paramname == "region" ~ "region",
                                 paramname == "state" ~ "state",
                                 paramname == "group_poll" ~ "pollutantGroup",
                                 paramname == "name_poll" ~ "pollutantName",
                                 paramname == "addparam" ~ "addressedParameters")
    
    # create variable for filter value
    param.filt <- eval(parse(text = param.name))
    
    # reformat param.filt if longer than one value
    if(length(param.filt) >1) {
      
      param.filt <- unlist(param.filt)
      
      # if the param is region, transform to numeric
      if(grepl("region", paramname, fixed = TRUE)) {
        
        param.filt <- as.numeric(param.filt)
      }
    }
    
    # do not filter df if param equals NULL
    if(is.null(eval(parse(text = param.name)))) {
      
      .data <- .data
      
    }
    
    # filter df if param does not equal NULL
    if(!is.null(eval(parse(text = param.name)))) {
      
      # params that can have multiple inputs
      mult.param <- c("region", "state", "group_poll", "name_poll", "addparam")
 
  # filter data
  .data <- .data %>%
  dplyr::filter(
    if(paramname == "year_min") {
      !!rlang::sym(tmdl.col) >= param.filt
    } else if (paramname == "year_max") {
      !!rlang::sym(tmdl.col) <= param.filt
    } else if(paramname %in% mult.param) {
      !!rlang::sym(tmdl.col) %in% param.filt
    }
  )
  }
  return(.data)
    }
}


# filter tmdl data
# this might be able to be rewritten with a purrr::map variant (more concise)
tmdl.df <- filt.df %>%
  paramCheck(paramname = "year_min") %>%
  paramCheck(paramname = "year_max") %>%
  paramCheck(paramname = "region") %>%
  paramCheck(paramname = "state") %>%
  paramCheck(paramname = "group_poll") %>%
  paramCheck(paramname = "name_poll") %>%
  paramCheck(paramname = "addparam")
  


# get count of TMDLs by unique combination of pollutant/action id/assessment unit
tmdl.actid.count <- dplyr::n_distinct(tmdl.df)

# get count of TMDLs by unique combination of pollutant/assessment unit
tmdl.noactid.count <- tmdl.df %>%
  dplyr::select(pollutant, assessmentUnitId) %>%
  dplyr::distinct() %>%
  dplyr::n_distinct()

```

In this filtered data set there are:

`r formatC(tmdl.actid.count, big.mark = ",")` unique combinations of actionId, assessmentUnitId and pollutant. `r formatC(tmdl.noactid.count, big.mark = ",")` unique combinations of assessmentUnitId and pollutant.

`r  if(params$filt.dt == TRUE){paste0("<h1>Filtered TMDL Results</h1>", "The data table below displays all TMDL records matching filters selected by the user. Each row represents a unique combination of pollutant, assessment unit, and action id.")}`


```{r filt.df.section, echo=FALSE}

# display filtered data table of tmdl results if user has selected it
if(params$filt.dt == TRUE){
  DT::datatable(tmdl.df)
}

```

`r  if(params$prod.output.plot == TRUE | prod.output.data == TRUE){paste0("<h1>TMDL Production History</h1>", "The data table below displays all TMDL records matching filters selected by the user. Each row represents a unique combination of pollutant, assessment unit, and action id.")}`

prod.output.plot: TRUE
  prod.output.data: TRUE